---
title: "README"
format: gfm
prefer-html: true
editor: visual
---

## Research Question

How does the opening of Houston's Red Line Light Rail correlate to changes in transportation choice and PM2.5 pollution?

## Research Context

-   Houston, Texas

    -   Red Line Light Rail

-   Timeline: 2000-2008

    -   The Red Line opened in 2004.

-   Hypothesis: Light rail openings increase pollution around stations,

    -   Reasoning: Light rail stations aggregate commuters throughout Houston, resulting in a concentration of pollution around these stations.

-   Confounding Factors

    -   Power Plants: Houston has several power plants that produce PM2.5

        -   The primary pollutant along the Red Line is the UH Central Power Station

    -   Roads/Highways/Intersections: Houston has some of the most congested roads in the U.S.

    -   Meteorological Factors: Various weather and climate influences on PM2.5 levels

        -   Temperature

        -   Wind

        -   Humidity

    -   Policies

        -   NAAQS

        -   TERP: Texas Emission Reduction Plan

### Power Plants

We used a Google API key to identify the coordinates for all power plants and stations along the Red Line.

-   Three main polluters along the Red Line

    -   UH Central Power Station: Primary polluter and provides energy to the University of Houston.

    -   Friendswood Energy Center: Supports the reliability of the local grid by generating energy during power fluctuations.

    -   Houston Holmes Ready-Mix Concrete Plant: A cement mixing plant that pollutes particulate matter.

```{r}
#| message: false
#| warning: false
library('knitr')
#| message: false
#| warning: false
library('tidyverse')
t <- read.csv('Untitled spreadsheet - Sheet1.csv')
kable(t)
```

### Red Line Stations

Below is a table of the 26 stations on Houston's Red Line

```{r}
t <- read.csv('Houston Data Collection.csv')
t <- t %>%
  select(Station, Opening.Dates, Address, Parking)
kable(t)
```

### Map of Houston

-   Red Line Stations: Black Points

-   Buffers: Red Circles

    -   Radius: 125 m

-   Highways: Yellow Lines

-   Power Plants: Blue Points

```{r}
#| warning: false
#| message: false

library('terra')
library('maptiles')
library('ggmap')

x <- vect('bg_x_vect.shp')
buff_sta <- vect('buff_sta.shp')
sta_pts <- vect('sta_pts.shp')
pow_pla_pts <- vect('pow_pla_pts.shp')
tsp <- vect('tsp.shp')

extent <- buffer(x, width = 10000)
bg <- get_tiles(ext(extent))



plot(bg)
points(sta_pts)
points(pow_pla_pts, col = 'blue')
lines(buff_sta, col = 'red')
lines(tsp, col = "yellow", lwd = 2)

```

## Linear Regression Model for PM2.5

We gathered PM2.5 data from NASA's Socioeconomic Data and Application Center. We combined the PM2.5 data with meteorological data from NASA's Global Land Data Assimilation System used use and DB-OLS model to identify connections. The regression model includes the relevant policies NAAQS and Texas Emissions Reduction Policy (TERP) which occurred during the time of our study. The model also includes the influence of metro stations opening with the MetroOpen variable and the influence of construction with the construction variable. Additionally we include time trend variables to account for continuous confounding factors. The model also includes the use of two dummy variables for construction and holiday days. By isolating these factors, the model serves to predict PM2.5 levels by each factor with a key focus on the influence of the MetroOpen variable.

```{r}
m <- read.csv('regression_model.csv')
m
```

## Spatial Data Mapping

### PM2.5 Data by PUMA

We identify pollution reductions per station by organizing stations into the PUMAs they reside in and identify the PM2.5 reduction by PUMAs. PUMAs are Public Use Microdata Areas provided by the U.S. census. Here we use PUMAS 4602, 4603, and 4604. We identify the pollution reduction by PUMA by creating a new linear regression model including the MetroOpen:as.factor(PUMACE10) as a variable.

```{r}
mPM2.5 <- read.csv('PM2.5PollutionByPumaRegressionModel.csv')
mPM2.5
```

```{r}
shapevect <- vect('shapevect.shp')
buff3 <- vect('buff3.shp')
buff2 <- vect('Buff2.shp')

osmpos <- create_provider(name = "CARTO.POSITRON",
                          url = "https://{s}.basemaps.cartocdn.com/light_all/{z}/{x}/{y}.png",
                          sub = c("a", "b", "c", "d"),
                          citation = "© OpenStreetMap contributors © CARTO ")
bg <- get_tiles(ext(buff3),provider = osmpos, crop = TRUE)
plot(bg, alpha=0.05)
plot(shapevect, add=TRUE)
plot(buff2, 
     "coef",
     type="interval",
     breaks=c(-34,-33.42,-30,-29.95, -23.94),
     col=map.pal("inferno"),
     cex.main=1.125,
     main="Average PM2.5 Change \n at Each Light Rail Station",
     plg=list( 
       title = "Change in PM2.5 \n (in Percents)",
       title.cex = 1, 
       cex = 1),
     add=TRUE)

```

### Transportation by PUMA

We identify the change in method of commute transportation by separating PUMA census data from the years 2004 and 2008 and tracking the percentage change for each of the three PUMAS. We tracked the percentage change in those who bike, walk, bus, or drive to work.

```{r}


pum_coef <- read.csv('pum_coef.csv', colClass=c("PUMACE10"="character"))
buff<-vect("buff_sta.shp")
shapevect<-vect('shapevect.shp')
shape_coef<-merge(shapevect, pum_coef, by="PUMACE10")

plot(bg, alpha=0.05)
plot(shapevect, add=TRUE)
plot(shape_coef,
     "change_bicycle",
     type="interval",
     breaks=c(-101,-100,-99.7,-99.5,-99.6),
     col=map.pal("reds"),
     plg=list( # parameters for drawing legend
       title = "Change in Bicycle Use \n (in Percents)",
       title.cex = 1, # Legend title size
       cex = 1),
     add=TRUE)
title("Average Bicycle Use Changes\n (in Percents) After Light Rail Openings", line = -20, adj=0.45,cex.main=1)

plot(buff, add=TRUE)

plot(bg, alpha=0.05)
plot(shapevect, add=TRUE)
plot(shape_coef,
     "change_walked",
     type="interval",
     breaks=c(-100, -99,-99.65,-98.5, -98),
     col=map.pal("grass"),
     plg=list( # parameters for drawing legend
       title = "Change in Walking Use \n (in Percents)",
       title.cex = 1, # Legend title size
       cex = 1),
     add=TRUE)
title("Average Walking Use Changes\n (in Percents) After Light Rail Openings", line = -20, adj=0.45,cex.main=1)
plot(buff, add=TRUE)

plot(bg, alpha=0.05)
plot(shapevect, add=TRUE)
plot(shape_coef, 
     "change_bus",
     type="interval",
     breaks=c(-100,-99,-98,-97.8,-97.5),
     col=map.pal("blues"),
     plg=list( # parameters for drawing legend
       title = "Change in Bus Use \n (in Percents)",
       title.cex = 1, # Legend title size
       cex = 1),
     add=TRUE)
title("Average Bus Use Changes\n (in Percents) After Light Rail Openings", line=-20, adj=0.45,cex.main=1)
plot(buff, add=TRUE)

plot(bg, alpha=0.05)
plot(shapevect, add=TRUE)
plot(shape_coef, 
     "change_car",
     type="interval",
     breaks=c(100,13,3.7,2.56,2.55),
     col=map.pal("oranges"),
     plg=list( # parameters for drawing legend
       title = "Change in Car Use \n (in Percents)",
       title.cex = 1, # Legend title size
       cex = 1),
     add=TRUE)
title("Average Car Use Changes\n (in Percents) After Light Rail Openings", line=-20, adj=0.45,cex.main=1)
plot(buff, add=TRUE)
```

## Conclusion

Light rail opening led to an average PM2.5 reduction of 12-29%. Areas with the most increase in car use saw the least reduction. After the 2004 opening of the red line, the use of buses, biking, and walking decreased between 98-100%, but car usage increased between 2-13% Stations with increased car use correlated with weaker reductions in PM2.5 levels compared to the overall reduction seen after the light rail opening

## References

### Data References

-   Di, Q., Y. Wei, A. Shtein, C. Hultquist, X. Xing, H. Amini, L. Shi, I. Kloog, R. Silvern, J. Kelly, M. B. Sabath, C. Choirat, P. Koutrakis, A. Lyapustin, Y. Wang, L. J. Mickley, and J. Schwartz. Daily and Annual PM2.5 Concentrations for the Contiguous United States, 1-km Grids, v1 (2000 - 2016). Palisades, New York: NASA Socioeconomic Data and Applications Center (SEDAC). Accessed 07/16/2024, <https://doi.org/10.7927/0rvr-4538>.

-   Di, Q., H. Amini, L. Shi, I. Kloog, R. Silvern, J. Kelly, M. B. Sabath, C. Choirat, P. Koutrakis, A. Lyapustin, Y. Wang, L. J. Mickley, and J. Schwartz. 2019. An Ensemble-based Model of PM2.5 Concentration Across the Contiguous United States with High Spatiotemporal Resolution. Environment International 130: 104909. Accessed 07/16/2024, <https://doi.org/10.1016/j.envint.2019.104909>.

-   Li, B., H. Beaudoing, and M. Rodell, NASA/GSFC/HSL (2020), GLDAS Catchment Land Surface Model L4 daily 0.25 x 0.25 degree GRACE-DA1 V2.2, Greenbelt, Maryland, USA, Goddard Earth Sciences Data and Information Services Center (GES DISC), Accessed: 07/16/2024, [10.5067/TXBMLX370XX8](https://doi.org/10.5067/TXBMLX370XX8)

### Policy References

-   Texas Commission on Environmental Quality. (2024, July 12). Texas Emissions Reduction Plan. Texas Commission on Environmental Quality. https://www.tceq.texas.gov/airquality/terp
