---
title: "README"
format: gfm
prefer-html: true
editor: visual
---

## Research Question

Which demographic groups are impacted by light rail openings in the United States.

## Research Context

-   Houston, Texas

    -   Red Line Light Rail

-   Timeline: 2000-2008

-   Hypothesis: Light rail openings increase pollution around stations

    -   Reasoning: Light rail stations aggregate commuters throughout Houston, resulting in a concentration of pollution around these stations.

    -   Confounding Factors

        -   Power Plants: Houston has several power plants that produce PM2.5

        -   Roads/Highways/Intersections: Houston has some of the most congested roads in the U.S.

        -   Meteorological Factors

    ### Table of Power Plants

```{r}
library('knitr')
t <- read.csv('Untitled spreadsheet - Sheet1.csv')
kable(t)
```

### Table of Red Line Stations

```{r}
t <- read.csv('Houston Data Collection - Charlotte.csv')
kable(t)
```

### Map of Houston

-   Red Line Stations: Black Points

-   Buffers: Red Circles

    -   Radius: 125 km

-   Highways: Yellow Lines

-   Power Plants: Blue Points

```{r}
#| warning: false
#| message: false

library('terra')
library('maptiles')
library('tidyverse')
library('ggmap')

x <- vect('bg_x_vect.shp')
buff_sta <- vect('buff_sta.shp')
sta_pts <- vect('sta_pts.shp')
pow_pla_pts <- vect('pow_pla_pts.shp')
tsp <- vect('tsp.shp')

extent <- buffer(x, width = 10000)
bg <- get_tiles(ext(extent))



plot(bg)
points(sta_pts)
points(pow_pla_pts, col = 'blue')
lines(buff_sta, col = 'red')
lines(tsp, col = "yellow", lwd = 2)

```

### Combining PM2.5 Data and Meteorological Data

```{r}
t <- read.csv('stations_data.csv')
head(t)
#kable(t)
```

### Linear Regression Model for PM2.5

```{r}
m <- read.csv('regression_model.csv')
m
```
