---
title: "README"
format: gfm
editor: visual
---

## Research Question

Which demographic groups are impacted by light rail openings in the United States.

## Research Context

-   Houston, Texas

    -   Red Line Light Rail

-   Timeline: 2000-2008

-   Hypothesis: Light rail openings increase pollution around stations

    -   Reasoning: Light rail stations aggregate commuters throughout Houston, resulting in a concentration of pollution around these stations.

    -   Confounding Factors

        -   Power Plants: Houston has several power plants that produce PM2.5

        -   Roads/Highways/Intersections: Houston has some of the most congested roads in the U.S.

        -   Meteorological Factors

-   Include table of confounding factor

```{r}
t <- read.csv('Untitled spreadsheet - Sheet1.csv')
kable(t)
```

-   Include table or map of stations

```{r}
library("knitr")
t <- read.csv('Houston Data Collection - Charlotte.csv')
kable(t)
```

```{r}
library("terra")

plot(bg)
points(pow_pla_pts, col = 'blue')
lines(tsp, col = 'yellow', lwd = 2)
lines(buff_sta, col = 'red')
```
